---
title: "Modelling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modelling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(individual)
library(ggplot2)
```

#Creating an epidemiological models

This article will walk you through the steps to create a basic SIR model, the
hello world of epidemiology. In doing so you will learn how to define, simulate,
analyse and make changes to your model.

###Defining the model structure

This part allows researchers to get a basic idea of what you are modelling. You
clearly define the states you are interested in modelling and initialise the
numbers of individuals beginning in each state.

Specifying your model also allows the “individual” package to build the data
structures to handle your simulation.

###Defining the processes

```{r}
population <- 1000
S <- State$new('S', population)
I <- State$new('I', 0)
R <- State$new('R', 0)
human <- Individual$new('human', list(S, I, R))
```

###Defining the processes

Now let's specify how individuals move between states.

In "individual", we can do that with "processes". They run every timestep and
specify model updates based on the current state of the model.

Process functions take a single argument, the [SimAPI](/reference/SimAPI.html').

Some of these processes can be repetitive, so it is often a good idea to
generate these functions, as demonstrated below:

```{r}
random_transition_generator <- function(from, to, rate) {
  function(api) {
    from_state <- api$get_state(human, from)
    StateUpdate$new(
      human,
      to,
      from_state[runif(length(from_state)) < rate]
    )
  }
}

processes <- list(
  random_transition_generator(S, I, .3),
  random_transition_generator(I, R, .5),
  random_transition_generator(R, S, .1)
)
```

Running the simulation

To see the model output, you simply run simulate. It will return a
[Simulation]('/reference/Simulation.html') object, which you can analyse
however you want.

```{r}
simulation <- simulate(human, processes, 100)

plot_states <- function(output) {
  # group the state counts into one column 
  cols <- colnames(output)[colnames(output) != 'timestep']
  reshaped <- reshape(
    output,
    varying = cols,
    idvar = 'timestep',
    timevar = 'state',
    direction = 'long',
    v.names = c('counts')
  )

  # make the state column more readable
  reshaped$state <- vapply(reshaped$state, function(i) cols[[i]], character(1))

  ggplot(
    reshaped,
    aes(x = timestep, y = counts, group = state)
  ) + geom_line(aes(color = state))
}

plot_states(simulation)
```

###Adding Variables

Quite often you will want to model continuous variables. These variables are
affected by the environment and can in turn affect state changes.

Let’s take immunity for example. We could decide that it should be acquired when
individuals individuals recover from infection and should reduce the chances of
immune individuals from becoming re-infected.

We would first want to update our model structure…

```{r}
# We can randomly assign immunity values between 0 and .2
immunity <- Variable$new('immunity', function(size) runif(size, 0, .2))
human <- Individual$new(
  'human',
  list(S, I, R),
  variables = list(immunity)
)
```
We can then start to describe how this affects our processes...

```{r}
infection <- function(api) {
  from_state <- api$get_state(human, S)
  # Get the immunity for susceptable humans and use the complement to modify the
  # infection rate
  rate_modifier <- 1 - api$get_variable(human, immunity)[from_state]
  rate <- .3
  StateUpdate$new(
    human,
    I,
    from_state[runif(length(from_state)) < (rate * rate_modifier)]
  )
}

recovery <- function(api) {
  from_state <- api$get_state(human, I)
  rate <- .2
  recovered <- from_state[runif(length(from_state)) < rate]
  list(
    StateUpdate$new(
      human,
      R,
      recovered
    ),
    VariableUpdate$new(
      human,
      immunity,
      api$get_parameters()$immunity_level,
      recovered
    )
  )
}

processes_with_immunity <- list(
  infection,
  random_transition_generator(I, R, .5),
  recovery
)
```

There are two new features in this example that have not yet been mentioned:

 1. Notice how the recovery process returns a list of updates. This allows it to
    boost immunity as well as change the same individuals’ states.
 2. We have used the parameters, object to specify the immunity level to set.
    This can be specified when we decide to run the simulation. 

You can see how this affects our outputs by re-simulating the model.

```{r}
simulation <- simulate(
  human,
  processes_with_immunity,
  100,
  parameters = list(immunity_level = .6)
)
plot_states(simulation)
```

You could also analyse the variables in your population.

Let’s say you wanted to see the average immunity in your population:

```{r}
simulation <- simulate(
  human,
  processes_with_immunity,
  100,
  parameters = list(immunity_level = .6),
  custom_renderers = list(
    function(api) {
      list(mean_immunity=mean(api$get_variable(human, immunity)))
    }
  )
)

ggplot(
  simulation,
  aes(x = timestep, y = mean_immunity)
) + geom_line()
```

And there you have it! We have created an SIR model with immunity and analysed
its outputs.
