---
title: "Performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(individual)
```

## Running big models

"individual" is designed for running big individual-based models. But if you find your model taking too long or consuming all of your memory, here are a few performance considerations:

###1. Use prefabs

Every time your processes ask for the state or a variable, there is an overhead associated with copying simulation data into R land. Imagine you have the model below:

```{r}
# Our model
population <- 100000
timesteps <- 100
S <- State$new('S', population/2)
I <- State$new('I', population/2)
R <- State$new('R', 0)
immunity <- Variable$new('immunity', function(size) runif(size, 0, 1))
human <- Individual$new('human', list(S, I, R), variables = list(immunity))

infection <- function(api) {
  from_state <- api$get_state(human, S)
  # Get the immunity for susceptable humans and use the complement to modify the
  # infection rate
  rate_modifier <- 1 - api$get_variable(human, immunity)[from_state]
  rate <- .3
  api$queue_state_update(
    human,
    I,
    from_state[runif(length(from_state)) < (rate * rate_modifier)]
  )
}

random_transition_generator <- function(from, to, rate) {
  function(api) {
    from_state <- api$get_state(human, from)
    api$queue_state_update(
      human,
      to,
      from_state[runif(length(from_state)) < rate]
    )
  }
}

render_state_sizes <- function(api) {
  api$render('infected_counts', length(api$get_state(human, I)))
  api$render('susceptable_counts', length(api$get_state(human, S)))
  api$render('recovered_counts', length(api$get_state(human, R)))
}

processes <- list(
  infection,
  random_transition_generator(I, R, .5),
  random_transition_generator(R, S, .1),
  render_state_sizes
)

system.time(simulate(human, processes, timesteps))
```

A lot of epidemiology models are quite similar. We've included some optimised, reusable processes and event listeners that provide significant speed improvements with very little effort:

```{r}
processes <- list(
  infection,
  individual::fixed_probability_state_change_process(human$name, I$name, R$name, .5),
  individual::fixed_probability_state_change_process(human$name, R$name, S$name, .1),
  individual::state_count_renderer_process(human$name, c(S$name, I$name, R$name))
)

system.time(simulate(human, processes, timesteps))
```

###2. Write your processes in C++

Unfortunately, we don't have a prefab for every situation. Please feel free to write one of your own!

These are the basic steps to add C++ processes to your R package:

1. Run `use_rcpp` to set your package up for C++ development
2. Add `individual` to the `LinkingTo` section of your package DESCRIPTION
3. Write your process

Processes in C++ are of type `Rcpp::XPtr<process_t>`. Once you've passed your process to R land, you can configure it as usual.

C++ processes program against the `ProcessAPI` class, which mirrors the R API (but is much more efficient). Below is an example of how you could translate the `infection` process:

```
#include <individual.h>
#include <Rcpp.h>

//[[Rcpp::export]]
Rcpp::XPtr<process_t> infection_cpp(
    const std::string human,
    const std::string S,
    const std::string I,
    const std::string immunity
    double rate
    ) {
    return Rcpp::XPtr<process_t>(
        new process_t([=] (ProcessAPI& api) {
              const auto& susceptable = api.get_state(human, S);
              const auto& immunity_values = api.get_variable(human, immunity);
              const auto uniform = Rcpp::runif();
              auto target = individual_index_t();
              for (auto s : susceptable) {
                if (uniform < ((1 - immunity_values[s]) * rate)) {
                  target.insert(s);
                }
              }
              api.queue_state_update(human, I, target);
        }),
        true
    );
}
```

The infection function is only a few extra lines and is likely to have a large speed improvement. Calls to `ProcessAPI` avoid copying any data by returning `const&` to simulation data structures.

You can pull your process into R land and configure your process like so:

```
processes <- list(
  infection_cpp(human$name, S$name, I$name, immunity$name, .3),
  random_transition_generator(I, R, .5),
  random_transition_generator(R, S, .1),
  render_state_sizes
)

simulate(human, processes, timesteps)
```

That's everything you need to scale your models up to millions of individuals!