---
title: "Scheduling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scheduling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(individual)
library(ggplot2)
library(reshape2)
```

## Adding scheduling to your model

This tutorial continues on from the "Modelling" vignette.

Sometimes want to model things that don't happen every timestep. Every change we've modelled so far happens at the end of the timestep. Here's a simple SIR model taken from the last tutorial:

```{r}
# Our model
population <- 1000
S <- State$new('S', population)
I <- State$new('I', 0)
R <- State$new('R', 0)
human <- Individual$new('human', list(S, I, R))

# Our processes
random_transition_generator <- function(from, to, rate) {
  function(api) {
    from_state <- api$get_state(human, from)
    api$queue_state_update(
      human,
      to,
      from_state[runif(length(from_state)) < rate]
    )
  }
}

render_state_sizes <- function(api) {
  api$render('susceptible_counts', length(api$get_state(human, S)))
  api$render('infected_counts', length(api$get_state(human, I)))
  api$render('recovered_counts', length(api$get_state(human, R)))
}

processes <- list(
  random_transition_generator(S, I, .3),
  random_transition_generator(I, R, .5),
  random_transition_generator(R, S, .1),
  render_state_sizes
)

# run the simulation
output <- simulate(
  human,
  processes,
  100
)

plot_states <- function(output) {
  ggplot(
    melt(output, 'timestep'),
    aes(x = timestep, y = value, group = variable)
  ) + geom_line(aes(color = variable))
}

# plot the output
plot_states(output)
```

At the moment, we model recovery by randomly selecting half of the infected population every timestep.

But what if we want to wait 5 timesteps for a recovery?

Enter events. Processes can schedule events to for a future timestep. The API also allows you to monitor individuals who are already scheduled to avoid rescheduling them.

Here's an example of how you could model recovery using scheduling.

```{r}
recovery_event <- Event$new('recovery')
recovery_event$add_listener(function(api, target) {
  api$queue_state_update(human, R, target)
})

recovery_process <- function(api) {
  infected <- api$get_state(human, I)
  already_scheduled <- api$get_scheduled(recovery_event)
  infected <- setdiff(infected, already_scheduled) 
  to_recover <- infected[runif(length(infected)) < .5]
  api$schedule(recovery_event, to_recover, 5)
}

processes <- list(
  random_transition_generator(S, I, .3),
  recovery_process,
  random_transition_generator(R, S, .1),
  render_state_sizes
)

# add the event to the human individual
human <- Individual$new('human', list(S, I, R), events = list(recovery_event))

# run the simulation
output <- simulate(
  human,
  processes,
  100
)

# plot the output
plot_states(output)
```

You could also schedule stochastic recovery times with a simple modification:

```{r}
recovery_process_stochastic <- function(api) {
  infected <- api$get_state(human, I)
  already_scheduled <- api$get_scheduled(recovery_event)
  infected <- setdiff(infected, already_scheduled) 
  to_recover <- infected[runif(length(infected)) < .5]
  # sample recovery times that are normally distributed around a mean of 5
  api$schedule(recovery_event, to_recover, pmax(rnorm(length(to_recover), 5), 0))
}

processes <- list(
  random_transition_generator(S, I, .3),
  recovery_process_stochastic,
  random_transition_generator(R, S, .1),
  render_state_sizes
)

# run the simulation
output <- simulate(
  human,
  processes,
  100
)

# plot the output
plot_states(output)
```

We have made two changes to this model:

1. We created a process called `recovery_process`. Instead of calling `api$queue_state_change` to perform a state change at the end of the timestep, it calls `api$schedule` to trigger an event in 5 steps time.
2. We created an event, `recovery_event` which we pass to the `simulate` function. You can add any number of listeners to this event to run when the event is triggered.

Voilà! We have added an event based process to our model.
